Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-03-31 talk/icooolps2009/paper.tex
finished the review of the paper until section 3.1 (excluded)

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}

@@

\numberofauthors{2}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{Softwaretechnik und Programmiersprachen}\\
       \affaddr{Institut für Informatik}\\ 
       \affaddr{Universitätsstra{\ss}e 1}\\
       \affaddr{D-40225 Düsseldorf}\\
       \affaddr{Deutschland}\\
       \email{cfbolz@gmx.de}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

@@

The PyPy project is trying to find approaches to generally ease the
implementation of dynamic languages. It started as a Python implementation in
Python, but has now extended its goals to be generally useful for implementing
other dynamic languages as well. The general approach is to implement an
interpreter for the language in a subset of Python. This subset is chosen in
such a way that programs in it can be compiled into various target environments,
such as C/Posix, the CLR or the JVM. The PyPy project is described in more
details in Section \ref{sect:pypy}.

@@

The PyPy project\footnote{http://codespeak.net/pypy} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython. RPython ("Restricted Python") is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLR and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

@@

The generation of loops works as follows: At first, everything is interpreted.
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

When a common loop is identified, the interpreter enters a
special mode (called tracing mode). When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

@@

It is important to understand when the tracer considers a loop in the trace to
be closed. This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: During
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code entered; during tracing
they are the place where the check for a closed loop is performed.

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand{\commentout}[1]{}

@@

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{Softwaretechnik und Programmiersprachen}\\
       \affaddr{Institut für Informatik}\\ 
       \affaddr{Universitätsstra{\ss}e 1}\\
       \affaddr{D-40225 Düsseldorf}\\
       \affaddr{Deutschland}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

@@

The PyPy project is trying to find approaches to generally ease the
implementation of dynamic languages. It started as a Python implementation in
Python, but has now extended its goals to be generally useful for implementing
other dynamic languages as well. The general approach is to implement an
interpreter for the language in a subset of Python. This subset is chosen in
such a way that programs in it can be compiled into various target environments,
such as C/Posix, the CLI or the JVM. The PyPy project is described in more
details in Section \ref{sect:pypy}.

@@

The PyPy project\footnote{http://codespeak.net/pypy} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{AACM-DLS07}. RPython ("Restricted Python") is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

\anto{XXX: are the following paragraphs really needed? I don't think this
  details are needed in Section \ref{sect:implementation}.  I would say
  something much shorter, see below.}

@@

\anto{By writing VMs in a high-level language, we keep the implementation of
  the language free of low-level details such as memory management strategy,
  threading model or object layout.  These features are automatically added
  during the translation process which consists in a series of steps, each
  step transforming the representation of the program produced by the previous
  one until we get the final executable.  As we will see later, this internal
  low-level representation of the program is also used as an input for the
  tracing JIT.}

@@

\sout{The generation of loops works as follows: at first, everything is interpreted.}
\anto{At first, when the program starts, everything is interpreted.}
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

When a \sout{common}\anto{hot} loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

\anto{XXX I think it's worth spending one more paragraph to explain what a
  trace really is, i.e. that it's a list of \textbf{sequential} operations,
  intermixed to guards which guarantee that this particular sequence is still
  valid.  At the moment, the definition of trace is not given explicitly and
  it's mixed with the details of how the JIT work}

@@

It is important to understand when the tracer considers a loop in the trace to
be closed. This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: during
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code entered; during tracing
they are the place where the check for a closed loop is performed.
